A comparison of design-based and model-based approaches for finite population spatial data -- Supplementary Material.

### 2021-10-21 inst/manuscript/manuscript.Rmd
minor updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

journal: "An awesome journal"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
output: rticles::elsevier_article
---

@@

```{r, results = "asis", eval = FALSE}
sim_one <- read_csv(here("inst", "output", "sim_one", "summ_output.csv"))
colnames(sim_one) <- c("Approach", "Bias", "RMSE", "MedAE", "Coverage", "PClose", "MedIL")
sim_one_table <- xtable(sim_one, digits = 4, caption = "Approach, mean bias (Bias), root-mean-squared error (RMSE), median absolute error (MedAE), 95 percent interval coverage (Coverage), proportion of times the approach was closer to the true value (PClose), and median interval length (MedIL)", type = "latex", latex.environments = "center", label = "tab:sim_one")
print(sim_one_table, include.rownames = FALSE, comment = FALSE)
```

@@

**Mike** For design-based, it's really RMSE -- how should we address this?
Figure \ref{fig:figeff} shows the relative efficiency of the four approaches from Table \ref{tab:designanalysis} with "IRS-Design" as the baseline:
\mbox{}
\begin{equation*}
\text{EFF} = \frac{\text{rMSPE of approach}}{\text{rMSPE of IRS-Design}},
\end{equation*}

where rMSPE is the root-mean-squared-prediction error. When there is no spatial correlation (top row), the four approaches have approximately equal rMSPE, even when the assumptions of the model-based approaches are violated. So, using GRTS or using a spatial model does not result in much, if any, loss in efficiency even if the response variable is not spatially correlated. When there is high spatial correlation (bottom row), the GRTS-Model approach tends to perform best, but difference in relative efficiency between GRTS-Model and GRTS-Design is small. In the lognormal, high partial sill settings (bottom-right facet), GRTS-Design outperforms IRS-Model by a large margin, suggesting that the design decision (whether to use IRS or GRTS) is more important than the analysis decision (whether to analyze using model assumptions or not).

@@

@@

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

journal: "TBD"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
output: rticles::elsevier_article
---

@@

@@

Figure \ref{fig:figeff} shows the relative efficiency of the four approaches from Table \ref{tab:designanalysis} with "IRS-Design" as the baseline:
\mbox{}
\begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

where rMS(P)E is the root-mean-squared error (design-based) or the root-mean-squared-prediction error (model-based). When there is no spatial correlation (top row), the four approaches have approximately equal rMSPE, even when the assumptions of the model-based approaches are violated. So, using GRTS or using a spatial model does not result in much, if any, loss in efficiency even if the response variable is not spatially correlated. When there is high spatial correlation (bottom row), the GRTS-Model approach tends to perform best, but difference in relative efficiency between GRTS-Model and GRTS-Design is small. In the lognormal, high partial sill settings (bottom-right facet), GRTS-Design outperforms IRS-Model by a large margin, suggesting that the design decision (whether to use IRS or GRTS) is more important than the analysis decision (whether to analyze using model assumptions or not).

@@

In addition to rMS(P)E and interval coverage, we also recorded average bias. The average bias is nearly zero for all approaches in all scenarios, so we omit a visualization of the results here. The supplementary material contains tables with mean bias, rMS(P)E, and interval coverage for all 36 simulation scenarios.

@@

# Data and Code Availability {.unnumbered}

This manuscript has a supplementary R package that contains all of the data and code used. Instructions for download at available at [https://github.com/michaeldumelle/DvMsp](https://github.com/michaeldumelle/DvMsp).

# Supplementary Material {.unnumbered}

In the supplementary material, we provide summary statistics for all 36 simulation scenarios.

### 2021-10-21 inst/manuscript/manuscript.tex
minor updates

  \journal{An awesome journal} % Sets Journal name

@@

  \title{A comparison of design-based and model-based approaches for finite
population spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matt Higham}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}

@@

\textbf{Mike} For design-based, it's really RMSE -- how should we
address this? Figure \ref{fig:figeff} shows the relative efficiency of
the four approaches from Table \ref{tab:designanalysis} with
``IRS-Design'' as the baseline: \mbox{} \begin{equation*}
\text{EFF} = \frac{\text{rMSPE of approach}}{\text{rMSPE of IRS-Design}},
\end{equation*}

where rMSPE is the root-mean-squared-prediction error. When there is no
spatial correlation (top row), the four approaches have approximately
equal rMSPE, even when the assumptions of the model-based approaches are
violated. So, using GRTS or using a spatial model does not result in
much, if any, loss in efficiency even if the response variable is not
spatially correlated. When there is high spatial correlation (bottom
row), the GRTS-Model approach tends to perform best, but difference in
relative efficiency between GRTS-Model and GRTS-Design is small. In the
lognormal, high partial sill settings (bottom-right facet), GRTS-Design
outperforms IRS-Model by a large margin, suggesting that the design
decision (whether to use IRS or GRTS) is more important than the
analysis decision (whether to analyze using model assumptions or not).

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which to select samples and analyze
data. The design-based approach incorporates randomness through sampling
to estimate a population parameter. The model-based approach
incorporates randomness through distributional assumptions to predict
the realized value of a random variable. Though these approaches have
often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
were particularly notable: the design decision (GRTS vs IRS) seems much
more important than the analysis decision (design-based vs model-based);
Independent of the analysis approach, there is no reason to prefer IRS
over GRTS for spatial data -- GRTS tends to perform at least as well as
IRS when there is no spatial correlation increasingly than IRS as the
strength of spatial correlation increases; the gap in relative
efficiency between GRTS-design and GRTS-model widens as the strength of
spatial correlation increases; and when the data are skewed, interval
coverage for all approaches improves both as the sample size increases
and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research. The
design-based approach relies on few assumptions, while the model-based
approach relies on rigid distributional ones. The Horvitz-Thompson
estimator of means and totals is unbiased and computationally efficient,
while model-based estimation of covariance parameters can be
computationally burdensome, especially for likelihood-based methods such
as REML that rely on inverting a covariance matrix. The design-based
approach also more naturally handles binary data, free from the more
complicated logistic regression formulation commonly used to handle
binary data in a model-based approach. The model-based approach,
however, can quantify the relationship between covariates (predictor
variables) and the response variable, something the design-based
approach cannot do naturally. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the process of study. Model selection is also possible using model-based
approaches and criteria such as likelihood ratio tests or AIC (Akaike,
1974). Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Model-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations. The benefits and drawbacks
of both approaches, alongside our theoretical and analytical
comparisons, should be heavily considered when choosing among them. This
is especially true from an analysis perspective, as we found that using
a spatially balanced sampling algorithm benefits both design-based and
model-based analyses.

>>>>

  \journal{TBD} % Sets Journal name

@@

  \title{A comparison of design-based and model-based approaches for finite
population spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matt Higham}

    \author[OSU]{Lisa Madsen}

@@

Figure \ref{fig:figeff} shows the relative efficiency of the four
approaches from Table \ref{tab:designanalysis} with ``IRS-Design'' as
the baseline: \mbox{} \begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

where rMS(P)E is the root-mean-squared error (design-based) or the
root-mean-squared-prediction error (model-based). When there is no
spatial correlation (top row), the four approaches have approximately
equal rMSPE, even when the assumptions of the model-based approaches are
violated. So, using GRTS or using a spatial model does not result in
much, if any, loss in efficiency even if the response variable is not
spatially correlated. When there is high spatial correlation (bottom
row), the GRTS-Model approach tends to perform best, but difference in
relative efficiency between GRTS-Model and GRTS-Design is small. In the
lognormal, high partial sill settings (bottom-right facet), GRTS-Design
outperforms IRS-Model by a large margin, suggesting that the design
decision (whether to use IRS or GRTS) is more important than the
analysis decision (whether to analyze using model assumptions or not).

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which to samples are selected and
data are analyzed. The design-based approach incorporates randomness
through sampling to estimate a population parameter. The model-based
approach incorporates randomness through distributional assumptions to
predict the realized value of a random variable. Though these approaches
have often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
are particularly notable: the design decision (GRTS vs IRS) seems much
more important than the analysis decision (design-based vs model-based);
independent of the analysis approach, there is no reason to prefer IRS
over GRTS for spatial data -- GRTS tends to perform at least as well as
IRS when there is no spatial correlation and increasingly better than
IRS as the strength of spatial correlation increases; the gap in
relative efficiency between GRTS-design and GRTS-model widens as the
strength of spatial correlation increases; and when the data are skewed,
interval coverage for all approaches improves both as the sample size
increases and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research.
Design-based approaches are often computationally efficient, while
model-based estimation of covariance parameters can be computationally
burdensome, especially for likelihood-based methods such as REML that
rely on inverting a covariance matrix. The design-based approach also
more naturally handles binary data, free from the more complicated
logistic regression formulation commonly used to handle binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable. The model-based approach also
yields estimated spatial covariance parameters, which help better
understand the process of study. Model selection is also possible using
model-based approaches and criteria such as likelihood ratio tests or
AIC (Akaike, 1974). Model-based approaches are capable of more efficient
small-area estimation than design-based approaches by leveraging
distributional assumptions in areas with few observed sites. Model-based
approaches can also compute site-by-site predictions at unobserved
locations and use them to construct informative visualizations. The
benefits and drawbacks of both approaches, alongside our theoretical and
analytical comparisons, should be heavily considered when choosing among
them. This is especially true from an analysis perspective, as we found
that using a spatially balanced sampling algorithm benefits both
design-based and model-based analyses.

\hypertarget{data-and-code-availability}{%
\section*{Data and Code Availability}\label{data-and-code-availability}}
\addcontentsline{toc}{section}{Data and Code Availability}

This manuscript has a supplementary R package that contains all of the
data and code used. Instructions for download at available at
\url{https://github.com/michaeldumelle/DvMsp}.

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

In the supplementary material, we provide summary statistics for all 36
simulation scenarios.

### 2021-10-21 inst/manuscript/supplementary.Rmd
minor updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data -- Supplementary Material.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
journal: "An awesome journal"
date: "`r Sys.Date()`"
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \renewenvironment{abstract}{}{}
output: rticles::elsevier_article
---

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data -- Supplementary Material.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
journal: "TBD"
date: "`r Sys.Date()`"
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \renewenvironment{abstract}{}{}
output: rticles::elsevier_article
---
